---
title: "reglScatterplot — Bioconductor integration tour"
author: "George Muñoz"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE
)
```

This notebook exercises every Bioconductor-class path the package supports.
Run it cell-by-cell in RStudio (Run > Run All) or in VSCode with the R
Jupyter kernel. All synthetic data — no downloads, no internet (apart from
the regl-scatterplot CDN dependency the widget always uses).

# Setup

```{r packages}
library(reglScatterplotR)
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(SummarizedExperiment)
    if (requireNamespace("SpatialExperiment", quietly = TRUE)) {
        library(SpatialExperiment)
    }
})
packageVersion("reglScatterplotR")
```

# Synthetic SingleCellExperiment

A small fake scRNA-seq object: 5 000 cells × 200 genes, with 8 clusters
arranged on a plausibly-clustered UMAP.

```{r sce-build}
set.seed(1L)
n_cells <- 5000L
n_genes <- 200L

counts <- matrix(rpois(n_genes * n_cells, lambda = 1.5),
                 nrow = n_genes, ncol = n_cells)
rownames(counts) <- paste0("Gene", seq_len(n_genes))
colnames(counts) <- paste0("Cell", seq_len(n_cells))

logc <- log2(counts + 1)

sce <- SingleCellExperiment(
    assays = list(counts = counts, logcounts = logc),
    colData = DataFrame(
        cluster = factor(sample(paste0("c", 1:8), n_cells, replace = TRUE)),
        depth   = colSums(counts),
        mito_pct = runif(n_cells, 0, 25)
    )
)

## Plausibly-clustered UMAP
mu <- matrix(rnorm(16, sd = 5), nrow = 8L, ncol = 2L)
clust_ix <- as.integer(sce$cluster)
umap <- mu[clust_ix, ] + matrix(rnorm(2L * n_cells, sd = 0.5), ncol = 2L)
reducedDim(sce, "UMAP") <- umap

sce
```

# Categorical colouring — cluster label

```{r sce-cluster}
reglScatterplot(sce,
                x = "UMAP",
                colorBy = "cluster",
                title = "UMAP coloured by cluster",
                height = 500)
```

# Continuous colouring — sequencing depth

`vmin = "p1"` / `vmax = "p99"` clips outliers so the palette gradient
remains useful when a few cells dominate.

```{r sce-depth}
reglScatterplot(sce,
                x = "UMAP",
                colorBy = "depth",
                continuousPalette = "magma",
                vmin = "p1", vmax = "p99",
                title = "Sequencing depth",
                height = 500)
```

# Gene-expression colouring

When `colorBy` matches a row name of the SCE, values are pulled from the
`logcounts` assay automatically. The `assay` argument can override which
assay to read from.

```{r sce-gene}
reglScatterplot(sce,
                x = "UMAP",
                colorBy = "Gene12",        # any rowname works
                continuousPalette = "viridis",
                title = "Gene12 (log-normalised)",
                height = 500)
```

# Combining categorical color + group filter

Pass `groupBy` to expose a second filter dimension in the legend.

```{r sce-group}
reglScatterplot(sce,
                x = "UMAP",
                colorBy = "cluster",
                groupBy = "cluster",  # could be a different colData column
                height = 500)
```

# SpatialExperiment — tissue coordinates

Skipped automatically when `SpatialExperiment` isn't installed.

```{r spe, eval = requireNamespace("SpatialExperiment", quietly = TRUE)}
spe <- SpatialExperiment(
    assays = list(counts = counts),
    colData = DataFrame(spot_type = sce$cluster),
    spatialCoords = umap            # stand-in for real tissue coords
)
reglScatterplot(spe,
                x = "spatial",
                colorBy = "spot_type",
                title = "Spatial map (SpatialExperiment)",
                height = 500)
```

# Custom palettes

```{r palettes}
mypal <- c("#1b9e77", "#d95f02", "#7570b3", "#e7298a",
           "#66a61e", "#e6ab02", "#a6761d", "#666666")
reglScatterplot(sce,
                x = "UMAP",
                colorBy = "cluster",
                customPalette = mypal,
                legendTitle = "Cluster",
                height = 500)
```

# Stress test — same SCE pipeline at higher density

```{r big-sce}
big_n <- 200000L
big_counts <- matrix(rpois(20L * big_n, lambda = 1), nrow = 20L)
rownames(big_counts) <- paste0("g", seq_len(20L))
big_logc <- log2(big_counts + 1)
big_sce <- SingleCellExperiment(
    assays = list(logcounts = big_logc),
    colData = DataFrame(
        cluster = factor(sample(letters[1:12], big_n, replace = TRUE))
    )
)
big_mu <- matrix(rnorm(24, sd = 6), 12L, 2L)
big_umap <- big_mu[as.integer(big_sce$cluster), ] +
    matrix(rnorm(2L * big_n, sd = 0.4), ncol = 2L)
reducedDim(big_sce, "UMAP") <- big_umap

reglScatterplot(big_sce,
                x = "UMAP",
                colorBy = "cluster",
                title = "200k cells",
                height = 500)
```

# Session info

```{r sessioninfo}
sessionInfo()
```
