---
title: "Scaling reglScatterplot to millions of points"
author: "George Muñoz"
date: "`r Sys.Date()`"
package: reglScatterplot
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Scaling reglScatterplot to millions of points}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

# What works at what scale

`reglScatterplot()` was designed for the size of typical single-cell and
spatial datasets, but it can push well past that.

| Point count       | Status       | Notes                                                         |
|------------------:|:-------------|:--------------------------------------------------------------|
| 1 - 500 000       | Flawless     | Below the auto performance-mode threshold; full interactivity |
| 500 k - 5 M       | Smooth       | `performanceMode` kicks in automatically                      |
| 5 M - 20 M        | Usable       | Use `pointSize = 1`, `opacity = 1`, drop `pointLabels`         |
| 20 M - 100 M      | Standalone HTML reaches RAM ceiling | Tile-based architectures (e.g. deepscatter) start to win |
| > 100 M           | Out of reach in-browser            | Server-side rendering / WebGPU territory                  |

# How the wire format works

To keep large datasets shippable inside a standalone `htmlwidget`, every
numeric channel is binary-encoded and base64-wrapped before transit:

| Channel                | Encoder                  | Precision        | Bytes / point |
|------------------------|--------------------------|------------------|--------------:|
| X / Y (normalised)     | `.toBase64U16()`         | 1 / 32 767       |             2 |
| Continuous color z     | `.toBase64U16Unit()`     | 1 / 65 535       |             2 |
| Categorical color z    | `.toBase64U16Int()`      | exact (< 65 536) |             2 |
| Filter ranges          | `toBase64()` (Float32)   | full f32         |             4 |

At 10 M points the resulting HTML file is around 80 - 90 MB - large but
finite. The same data with Float32 everywhere would be ~150 MB.

# A benchmark you can run yourself

```{r bench}
library(reglScatterplot)

bench_sizes <- c(1e4, 1e5, 1e6, 5e6)
for (n in bench_sizes) {
    df <- data.frame(x = rnorm(n), y = rnorm(n), v = runif(n))
    t0 <- Sys.time()
    w  <- reglScatterplot(df, x = "x", y = "y", colorBy = "v",
                          height = 600)
    payload <- htmlwidgets:::toJSON(w$x)
    cat(sprintf("n = %s : build = %.2fs, payload = %.1f MB\n",
                format(n, big.mark = ","),
                as.numeric(Sys.time() - t0, units = "secs"),
                nchar(payload) / 1024 / 1024))
    rm(df, w, payload); gc(verbose = FALSE)
}
```

On a 2020-era laptop with an RTX 2060, 5 M points takes ~1.5s on the R side
and another ~2s for the browser to parse and upload to the GPU; pan/zoom
then runs at 60 fps.

# Sizing inside the host viewport

The widget honours an explicit pixel `height` verbatim. If the value
exceeds the height of the host window (small browser tab, RStudio Viewer
in a tiling WM, narrow Jupyter notebook column, etc.), the bottom of
the canvas is clipped by the host - not by `reglScatterplot`.

```{r sizing, eval = FALSE}
# Bad in small viewports: a 500 px tall widget overflows a 450 px window.
reglScatterplot(df, x = "x", y = "y", height = 500)

# Good: fill whatever vertical space is available.
reglScatterplot(df, x = "x", y = "y", height = "100%")

# Also good: omit `height` entirely - the sizingPolicy fills the viewer pane.
reglScatterplot(df, x = "x", y = "y")
```

Knitting to HTML produces a full-page document where the widget can take
as much height as you give it, so the same code that clips in the Viewer
pane prints cleanly in a knit report. This is purely a viewport effect.

# Memory levers for very large data

When you really want to push past 5 M, every per-point byte counts.
Suggested defaults for huge inputs:

```{r huge, eval = FALSE}
reglScatterplot(huge_df,
    x = "x", y = "y",
    pointSize = 1,            # one pixel per point
    opacity = 1,              # no blending math
    showAxes = FALSE,         # drops the D3 axis layer
    showTooltip = FALSE,      # frees per-point hit-test work
    enableDownload = FALSE,   # no html2canvas / jsPDF download
    pointLabels = NULL        # don't ship gene names
)
```

Things you might think help but don't:
* Reducing `vmin` / `vmax` clip range - colour scale only, not memory.
* Setting `legendPosition = "bottom-left"` - cosmetic, no perf impact.

# Comparison with other R packages

`reglScatterplot` is one of three credible options for high-volume scatter
in R. They aren't doing the same thing:

| Package         | Interactive? | Best at                              | Limit                        |
|-----------------|:------------:|--------------------------------------|------------------------------|
| `reglScatterplot` | Yes        | 1 - 20 M points in HTML / Shiny       | Browser RAM / VRAM           |
| `plotly` (+ `toWebGL()`) | Yes | < 500 k points, broad feature set    | JSON payload bloats past 1 M |
| `scattermore`   | No (static)  | Quickly rasterising 10 M+ to a PNG    | No pan / zoom interactivity  |
| `ggplot2`       | No (static)  | Publication graphics, small data      | Practical ceiling ~50 k pts  |

The right choice depends on what you need:

* Want a printable figure? `ggplot2` or `scattermore`.
* Want to embed an interactive plot in an HTML report? `reglScatterplot`.
* Need brush, click, and faceted layouts more than scale? `plotly`.

# Where the next jump comes from

For genuinely huge data (multi-modal CosMx slides, whole-atlas integrations
beyond ~50 M cells), no in-browser library is the right answer today. The
viable paths are:

1. **Tile-based architectures** - precompute spatial tiles on disk, only
   load what the viewport needs. See `deepscatter` (Apache Arrow + Parquet
   tiles). Requires a server or a static tile directory.
2. **Server-side rendering** - send camera state to a Python / Julia
   backend that renders frames; stream them as images. Lower fidelity but
   independent of the client.
3. **WebGPU** - browser support is maturing; offers compute shaders that
   would let us do GPU-side filtering and density binning. Currently a
   two-year horizon.

For now, `reglScatterplot` covers the typical single-cell, spatial and
fold-change use cases comfortably. If you find yourself loading the same
10 M+ dataset repeatedly, the right next step is to switch to a tile
server, not a faster scatterplot.

# Session info

```{r sessionInfo, eval = TRUE}
sessionInfo()
```
