---
title: "reglScatterplotR — full feature tour"
author: "reglScatterplotR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(reglScatterplotR)
```

> **How to use this notebook.** Run it chunk-by-chunk in RStudio (each plot
> appears in the Viewer / inline) or knit the whole thing to a self-contained
> HTML report. Every browser dependency is bundled, so it works **offline**.
> Chunks that need an optional package (Seurat, AnnData, SCE, …) skip
> gracefully with a message if it is not installed.

# 0. The bundled dataset

A 3,400-cell synthetic single-cell embedding ships with the package — no
download needed.

```{r dataset}
data(reglScatterExample)
str(reglScatterExample)
```

# 1. Basics

## 1.1 Minimal plot (data.frame)

```{r basic}
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2")
```

## 1.2 Colour by a categorical column

```{r color-categorical}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2",
  colorBy = "celltype",
  legendTitle = "Cell type"
)
```

## 1.3 Colour by a continuous column

```{r color-continuous}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2",
  colorBy = "CD3D",
  continuousPalette = "viridis",
  legendTitle = "CD3D"
)
```

## 1.4 Fixed point colour (no mapping)

```{r fixed-color}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2",
  pointColor = "#0072B2"
)
```

# 2. Point appearance

## 2.1 Point size & opacity

```{r size-opacity}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  pointSize = 6, opacity = 0.5
)
```

## 2.2 `pixelRatio` — crisp output in the RStudio Viewer

The Viewer reports `devicePixelRatio = 1` on HiDPI screens, so the default now
super-samples (`max(devicePixelRatio, 2)`). Override for memory or extra
sharpness:

```{r pixelratio}
# Extra-sharp (e.g. for screenshots / export)
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  pixelRatio = 3
)
```

# 3. Palettes

## 3.1 Built-in categorical / continuous palettes

```{r palettes-builtin}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "percentMito",
  continuousPalette = "magma", legendTitle = "% mito"
)
```

## 3.2 Custom colours for categories

```{r palette-custom}
celltypes <- levels(reglScatterExample$celltype)
my_cols <- setNames(
  grDevices::hcl.colors(length(celltypes), "Dark 3"),
  celltypes
)
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  customColors = my_cols
)
```

# 4. Filtering

## 4.1 Per-variable filter sliders (`filterBy`)

`filterBy` takes a data.frame of numeric columns that become interactive
range filters in the widget.

```{r filter}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  filterBy = reglScatterExample[, c("nCount", "nFeature", "percentMito")]
)
```

# 5. Axes, ranges & framing

## 5.1 Titles, axis labels, tooltip

```{r labels}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  title = "PBMC — UMAP", xlab = "UMAP 1", ylab = "UMAP 2",
  showAxes = TRUE, showTooltip = TRUE
)
```

## 5.2 Manual ranges, padding, auto-fit

```{r ranges}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  rangePadding = 0.3, autoFit = TRUE
)
```

## 5.3 Continuous colour scaling: `vmin`/`vmax`, `centerZero`

```{r vmin-vmax}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "CD3D",
  vmin = 0, vmax = 4, continuousPalette = "viridis"
)
```

# 6. Theming the canvas & legend

The legend is a frosted-glass card; tune its translucency with `legendOpacity`
(0–1) and `legendBlur` (px).

```{r theme}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  backgroundColor = "#111418", axisColor = "#cccccc",
  legendBg = "#1c2128", legendText = "#e6edf3",
  legendOpacity = 0.6, legendBlur = 12,
  legendPosition = "bottom-left", draggableLegend = TRUE,
  fontSize = 13, legendFontSize = 12
)
```

# 7. Toolbar & encodings

## 7.1 In-plot toolbar

`toolbarPosition` adds pan / lasso / zoom-to-selection / reset / screenshot
buttons (`"left"` vertical, `"top"` horizontal, `"none"` to hide).

```{r toolbar}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  toolbarPosition = "left"
)
```

## 7.2 Zoom to the lasso selection

```{r zoom-on-selection}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  toolbarPosition = "left", zoomOnSelection = TRUE
)
```

> **Reading the lasso selection in R.** The selection is reported to Shiny as
> `input$<plotId>_selected` (see section 11). Unlike the Python package's
> `w.selection`, a *static* R Markdown has no channel to send it back to R — use
> Shiny (or the `sp-selection` DOM event from JS) for that round-trip.

## 7.3 Encode size / opacity by data

Map a numeric variable to point size or opacity (they share one channel).

```{r size-opacity-by}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype", sizeBy = "nCount")
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype", opacityBy = "CD3D")
```

# 8. Export (PNG / SVG / PDF)

`enableDownload = TRUE` adds an in-plot export button (also on the toolbar).

```{r export}
reglScatterplot(reglScatterExample,
  x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  enableDownload = TRUE
)
```

# 9. Other input objects

The same `reglScatterplot()` call dispatches on the object class.

## 9.1 Plain numeric matrix

```{r matrix}
m <- as.matrix(reglScatterExample[, c("UMAP_1", "UMAP_2")])
reglScatterplot(m, colorBy = reglScatterExample$celltype)
```

## 9.2 SingleCellExperiment

```{r sce}
if (requireNamespace("SingleCellExperiment", quietly = TRUE)) {
  set.seed(1)
  n <- 500
  sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(logcounts = matrix(rnorm(10 * n), nrow = 10,
                  dimnames = list(paste0("g", 1:10), NULL))),
    colData = S4Vectors::DataFrame(
      cluster = factor(sample(letters[1:4], n, replace = TRUE)))
  )
  SingleCellExperiment::reducedDim(sce, "UMAP") <- matrix(rnorm(2 * n), ncol = 2)
  reglScatterplot(sce, x = "UMAP", colorBy = "cluster")
} else {
  message("SingleCellExperiment not installed — skipping.")
}
```

## 9.3 SpatialExperiment (spatial coords)

```{r spe}
if (requireNamespace("SpatialExperiment", quietly = TRUE)) {
  set.seed(2)
  n <- 400
  spe <- SpatialExperiment::SpatialExperiment(
    assays = list(counts = matrix(rpois(5 * n, 3), nrow = 5)),
    spatialCoords = matrix(runif(2 * n, 0, 100), ncol = 2,
                           dimnames = list(NULL, c("x", "y"))),
    colData = S4Vectors::DataFrame(region = factor(sample(1:3, n, TRUE)))
  )
  reglScatterplot(spe, x = "spatial", colorBy = "region")
} else {
  message("SpatialExperiment not installed — skipping.")
}
```

## 9.4 Seurat

```{r seurat}
if (requireNamespace("SeuratObject", quietly = TRUE)) {
  set.seed(3)
  n <- 400
  counts <- matrix(rpois(20 * n, 5), nrow = 20,
                   dimnames = list(paste0("Gene", 1:20), paste0("Cell", 1:n)))
  so <- SeuratObject::CreateSeuratObject(counts = counts)
  so$celltype <- factor(sample(c("A", "B", "C"), n, replace = TRUE))
  emb <- matrix(rnorm(2 * n), ncol = 2,
                dimnames = list(colnames(so), c("UMAP_1", "UMAP_2")))
  so[["umap"]] <- SeuratObject::CreateDimReducObject(
    embeddings = emb, key = "UMAP_", assay = "RNA")
  reglScatterplot(so, x = "umap", colorBy = "celltype")
} else {
  message("SeuratObject not installed — skipping.")
}
```

## 9.5 AnnData (anndataR)

```{r anndata}
if (requireNamespace("anndataR", quietly = TRUE)) {
  set.seed(4)
  n <- 400; g <- 15
  ad <- anndataR::AnnData(
    X = matrix(rpois(n * g, 2), nrow = n,
               dimnames = list(NULL, paste0("Gene", 1:g))),
    obs = data.frame(celltype = factor(sample(c("X", "Y", "Z"), n, TRUE))),
    obsm = list(X_umap = matrix(rnorm(2 * n), ncol = 2))
  )
  reglScatterplot(ad, x = "UMAP", colorBy = "celltype")
} else {
  message("anndataR not installed — skipping.")
}
```

# 10. Synchronised plots

Plots sharing a `syncPlots` group link their pan/zoom, lasso **selection** and
legend/range **filtering** — drag or hide a category in one and the others
follow. Lay them out with `htmltools` (this is the R counterpart of
`reglscatterpy`'s `compose()`).

```{r sync, eval=FALSE}
p1 <- reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2",
  colorBy = "celltype", plotId = "p1", syncPlots = c("p1", "p2"))
p2 <- reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2",
  colorBy = "CD3D", plotId = "p2", syncPlots = c("p1", "p2"))
htmltools::tagList(
  htmltools::div(style = "display:flex; gap:8px; height:460px",
    htmltools::div(style = "flex:1", p1),
    htmltools::div(style = "flex:1", p2)))
```

# 11. Shiny

In Shiny use `reglScatterplotOutput()` / `renderReglScatterplot()`, and
`enableReglScatterplotSync()` / `updateReglPointSize()` to drive plots from the
server. A lasso selection is reported to `input$<plotId>_selected`
(`$indices`, `$count`), so you can react to it server-side. See
`?reglScatterplotOutput`.

```{r shiny-selection, eval=FALSE}
# observeEvent(input$plot_selected, {
#   sel <- input$plot_selected$indices      # selected row indices
#   message(length(sel), " cells selected")
# })
```

```{r shiny, eval=FALSE}
library(shiny)
ui <- fluidPage(reglScatterplotOutput("plot", height = "600px"))
server <- function(input, output, session) {
  output$plot <- renderReglScatterplot({
    reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2",
      colorBy = "celltype", plotId = "plot")
  })
}
shinyApp(ui, server)
```

# Session info

```{r sessioninfo}
sessionInfo()
```
