---
title: "reglScatterplotR — interactive WebGL scatterplots"
author: "reglScatterplotR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(reglScatterplotR)
# Dev tip: to preview un-installed local changes without reinstalling each time,
# replace the line above with (adjust the path):
#   devtools::load_all("~/Desktop/reglScatterplotR")
```

> A guided tour. Run it chunk-by-chunk in RStudio (each plot shows in the Viewer)
> or knit to a self-contained HTML report — every browser dependency is bundled,
> so it works **offline**. Chunks needing an optional package (Seurat, AnnData,
> SCE, …) skip gracefully if it isn't installed. This is the R companion to the
> Python package [`reglscatterpy`](https://github.com/george123ya/reglscatterpy);
> both drive the same widget.

# Setup — the bundled dataset

A 3,400-cell synthetic single-cell `data.frame` ships with the package (UMAP
coords, a `celltype` factor, QC metrics and a `CD3D` marker), so you can plot
immediately.

```{r data}
data(reglScatterExample)
str(reglScatterExample)
```

# 1. Basics

```{r basic}
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2", colorBy = "celltype")
```

# 2. Colour & palettes

Categorical (ColorBrewer) and continuous (viridis) palettes; clip with
`vmin`/`vmax`; supply your own with `customColors`.

```{r color}
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2",
  colorBy = "CD3D", continuousPalette = "viridis", vmax = "p99")

cols <- setNames(grDevices::hcl.colors(8, "Dark 3"),
                 levels(reglScatterExample$celltype))
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2",
  colorBy = "celltype", customColors = cols)
```

# 3. Encode size / opacity by data

```{r encodings}
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2",
  colorBy = "celltype", sizeBy = "nCount")
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2",
  colorBy = "celltype", opacityBy = "CD3D")
```

# 4. Styling & legend

The legend is a frosted card — tune `legendOpacity` / `legendBlur`; its title
defaults to the colour variable name.

```{r styling}
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  title = "PBMC — UMAP", backgroundColor = "#0d1117", axisColor = "#cccccc",
  legendBg = "#1c2128", legendText = "#e6edf3", legendOpacity = 0.6,
  legendBlur = 12, legendPosition = "bottom-left")
```

# 5. Toolbar & zoom-on-selection

`toolbarPosition` adds pan / lasso / zoom-to-selection / reset / screenshot
(drag the grip to move it). `zoomOnSelection = TRUE` auto-frames a lasso.

```{r toolbar}
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  toolbarPosition = "left", zoomOnSelection = TRUE)
```

# 6. Filtering (distribution sliders)

`filterBy` adds a panel of density sliders — drag to keep cells within ranges,
live.

```{r filter}
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  filterBy = reglScatterExample[, c("nCount", "nFeature", "percentMito")])
```

# 7. Richer tooltips

```{r tooltips}
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  tooltipBy = c("nCount", "percentMito"))
```

# 8. Synced & linked plots

Plots sharing a `syncPlots` group link their pan/zoom, lasso selection and
legend/range filtering. Lay them out with `htmltools` (the R counterpart of
`reglscatterpy`'s `compose()`).

```{r sync, eval=FALSE}
p1 <- reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2",
  colorBy = "celltype", plotId = "p1", syncPlots = c("p1", "p2"))
p2 <- reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2",
  colorBy = "CD3D", plotId = "p2", syncPlots = c("p1", "p2"))
htmltools::tagList(htmltools::div(style = "display:flex; gap:8px; height:460px",
  htmltools::div(style = "flex:1", p1), htmltools::div(style = "flex:1", p2)))
```

`crosstalk` links the lasso selection + filtering with other widgets (DT,
plotly, leaflet) — no Shiny needed:

```{r crosstalk, eval=FALSE}
library(crosstalk)
sd <- SharedData$new(reglScatterExample)
bscols(
  reglScatterplot(sd, x = "UMAP_1", y = "UMAP_2", colorBy = "celltype"),
  DT::datatable(sd))   # lasso the plot -> the table filters to those cells
```

# 9. Other input objects

The same `reglScatterplot()` call dispatches on the object class.

```{r matrix}
m <- as.matrix(reglScatterExample[, c("UMAP_1", "UMAP_2")])
reglScatterplot(m, colorBy = reglScatterExample$celltype)   # plain matrix
```

```{r sce}
if (requireNamespace("SingleCellExperiment", quietly = TRUE)) {
  set.seed(1); n <- 500
  sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(logcounts = matrix(rnorm(10 * n), nrow = 10,
                  dimnames = list(paste0("g", 1:10), NULL))),
    colData = S4Vectors::DataFrame(cluster = factor(sample(letters[1:4], n, TRUE))))
  SingleCellExperiment::reducedDim(sce, "UMAP") <- matrix(rnorm(2 * n), ncol = 2)
  reglScatterplot(sce, x = "UMAP", colorBy = "cluster")
} else message("SingleCellExperiment not installed — skipping.")
```

```{r seurat}
if (requireNamespace("SeuratObject", quietly = TRUE)) {
  set.seed(3); n <- 400
  counts <- matrix(rpois(20 * n, 5), nrow = 20,
                   dimnames = list(paste0("Gene", 1:20), paste0("Cell", 1:n)))
  so <- suppressWarnings(SeuratObject::CreateSeuratObject(counts = counts))
  so$celltype <- factor(sample(c("A", "B", "C"), n, replace = TRUE))
  emb <- matrix(rnorm(2 * n), ncol = 2,
                dimnames = list(colnames(so), c("UMAP_1", "UMAP_2")))
  so[["umap"]] <- SeuratObject::CreateDimReducObject(emb, key = "UMAP_", assay = "RNA")
  reglScatterplot(so, x = "umap", colorBy = "celltype")
} else message("SeuratObject not installed — skipping.")
```

```{r anndata}
if (requireNamespace("anndataR", quietly = TRUE)) {
  set.seed(4); n <- 400; g <- 15
  adn <- anndataR::AnnData(
    X = matrix(rpois(n * g, 2), nrow = n, dimnames = list(NULL, paste0("Gene", 1:g))),
    obs = data.frame(celltype = factor(sample(c("X", "Y", "Z"), n, TRUE))),
    obsm = list(X_umap = matrix(rnorm(2 * n), ncol = 2)))
  reglScatterplot(adn, x = "UMAP", colorBy = "celltype")
} else message("anndataR not installed — skipping.")
```

`SpatialExperiment` (use `x = "spatial"`) and `Monocle3` `cell_data_set` (an SCE
subclass) work the same way.

# 10. Export

`enableDownload = TRUE` adds a PNG/SVG/PDF button (also on the toolbar).

```{r export}
reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2", colorBy = "celltype",
  enableDownload = TRUE)
```

# 11. Shiny — selection & single-cell workflows

In Shiny use `reglScatterplotOutput()` / `renderReglScatterplot()`. A lasso
selection is reported to `input$<plotId>_selected` (`$indices`, `$count`) — this
is the R selection round-trip (a static R Markdown has no back-channel, unlike
Python's `w.selection`).

```{r shiny, eval=FALSE}
library(shiny)
ui <- fluidPage(reglScatterplotOutput("plot", height = "600px"),
                verbatimTextOutput("info"))
server <- function(input, output, session) {
  output$plot <- renderReglScatterplot({
    reglScatterplot(reglScatterExample, x = "UMAP_1", y = "UMAP_2",
      colorBy = "celltype", plotId = "plot", toolbarPosition = "left")
  })
  output$info <- renderPrint({
    sel <- as.integer(unlist(input$plot_selected$indices))  # JS sends a list -> coerce
    if (length(sel)) {
      sub <- reglScatterExample[sel + 1L, ]         # 0-based JS -> 1-based R
      table(sub$celltype)                           # composition; or run
    }                                               # scran/Seurat DE here
  })
}
shinyApp(ui, server)
```

The Python package surfaces these as convenience methods on the selection
(`w.subset()`, `w.annotate()`, `w.composition()`, `w.diff_expression()`); in R
run the equivalent on `input$<id>_selected$indices` server-side.

# Session info

```{r sessioninfo}
sessionInfo()
```
